Distinguishing Spin-Aligned and Isotropic Black Hole Populations With Gravitational Waves

### 2017-04-12 paper/AlignedIsotropicGW.tex
Fix #6: chi, chieff, a confusion.

Figure \ref{fig:O1-posteriors} shows an approximation to the posterior
inferred on $\chieff$ for the three likely \ac{GW} detections
GW150914, GW151226 and LVT151012 from \citet{O1-BBH}.  Because samples
drawn from the posterior on $\chieff$ are not publicly released at
this time, we have approximated the posterior as a Gaussian
distribution with the same mean and 90\% credible interval as quoted
in \citet{O1-BBH}.  There is essentially no posterior support for
$\chi \gtrsim 0.5$ as inferred for the majority of the
electromagnetically-detected stellar-mass black hole population (see
Section \ref{sec:discussion}).  \ilya{[Figure 1 shows an absence of
  support for $\chieff \gtrsim 0.5$.  It does not directly say
  anything about $\chi$.  In fact, Figure 3 makes it quite clear that
  $\chi$ can have support beyond 0.5 as long as spins are isotropic.
  Meanwhile, the EM observations describe $\chi$, not $\chieff$, since
  only one component is a BH in those X-ray binaries.  In a way, this
  whole paper is about the difference between $\chi$ and $\chieff$
  (depending on the mis-alignment angles), so conflating them here and
  in the caption of Figure 1 is very confusing.]}  The analysis here
is relatively insensitive to the precise details of the posterior
distributions; other conclusions are more sensitive.  In particular,
our Gaussian approximation does permit $\chieff = 0$ for GW151226
while the true posterior rules this out at high confidence
\citep{2016PhRvL.116x1103A,O1-BBH}.

\begin{figure}
  \plotone{../plots/chi-eff-mock-posteriors}
  \caption{\label{fig:O1-posteriors} Approximate posteriors on
    $\chieff$ from the LIGO O1 observations in \citet{O1-BBH}.  We
    approximate the posteriors reported in \citet{O1-BBH} using
    Gaussians with the same median and 90\% credible interval as
    reported in \citet{O1-BBH}.  It is notable that none of the
    $\chieff$ posteriors extends to the high spin magnitudes inferred
    from several electromagnetically-detected stellar-mass black hole
    systems \ilya{[Again, $\chi$ and $\chieff$ are being conflated; see comment on 2nd paragraph of intro.]} (see \citet{2015PhR...548....1M} for a summary of such
    measurements).}
\end{figure}

Small values of $\chieff$ as exhibited in these systems can result
from either intrinsically small spins or larger spins whose direction
is mis-aligned with the orbital \sout{plane} \ilya{angular momentum} of the binary (i.e.\ spin
vectors with small $z$-components).  Mis-alignment is capable of
producing \emph{negative} values of $\chieff$, however, whereas aligned
spins will always have $\chieff \geq 0$.  This difference provides
strong discriminating power between the two angular distributions,
even without good information about the magnitude distribution; to the
extent that data favour negative $\chieff$ they weigh heavily against
aligned models.  To quantify the degree of support for these two
alternate explanations of small $\chieff$ values in the merging
\ac{BBH} population, we compared the Bayesian evidence for various
simple models of the spin population using the \ac{GW} data set.

Each of our models for the merging \ac{BBH} spin population assumes
that the merging black holes are of equal mass (this is marginally
consistent with the three observations \citep{O1-BBH}, and the
$\chieff$ distribution is not particularly sensitive to the mass ratio
between the merging objects---see Section \ref{sec:mass-ratio}).  We
assume that the population spin distribution factorises into a
distribution for the spin magnitude \ilya{$\chi$} and a distribution for the
spin angles.  Finally, we assume that the distribution of spins is
common to each component in a merging binary.  Choosing one of three
magnitude distributions
\begin{equation}
  \label{eq:magnitude-dists}
  p(\ilya{\chi}) = \begin{cases}
    2\left(1-\chi \right) & \textnormal{``low''} \\
    1 & \textnormal{``flat''} \\
    2 \chi & \textnormal{``high''}
  \end{cases},
\end{equation}
and pairing with either an isotropic angular distribution or a
distribution that generates perfect alignment with the positive $z$
axis yields six different models for the $\chieff$ distribution.
These models are shown in Figure \ref{fig:chieff-distribution-models}.

@@

The distributions in Eq.\ \eqref{eq:magnitude-dists} are not meant to
represent any particular physical model, but rather to capture our
uncertainty about the spin magnitude distribution, discussed in detail
in Section \ref{sec:discussion}; neither observations nor population
synthesis codes can at this point authoritatively suggest {\it any}
particular spin distribution \citep{2015PhR...548....1M}.  Our models,
however, allow us to see how sensitive the $\chieff$ distribution is
to spin alignment given uncertainties about the spin magnitudes.

@@

Estimates of the rate of \ac{BBH} coalescences give a reasonable
chance of 10 additional \ac{BBH} detections in the next two years
\citep{O1-BBH,2016ApJ...833L...1A,2016ApJS..227...14A}.  Assuming \ilya{10} \sout{an}
additional \sout{10} detections with similar observational uncertainties \ilya{[I think it would be useful to specify (in an appendix?) what model we use for the observational uncertainty on $\chieff$.]}
drawn from each of our six zero-parameter models for the spin
distribution in addition to the three existing detections from O1, we
find the odds ratios shown in Figure \ref{fig:O2-predictions}.  We
find that most scenarios with an additional 10 detections allow the
\sout{correct} \ilya{simulated} angular distribution to be \sout{distinguished} \ilya{inferred} with greater than
$5\sigma$ ($2.9 \times 10^{-7}$ odds) credibility; and in the most
pessimistic case the distinction is typically
$\OTwoSigmaIsoAlignedMin\sigma$ ($\OTwoOddsIsoAlignedMin$ odds ratio).
While such future detections should permit a confident distinction
between \emph{angular} distributions, we would remain much less certain
about the \emph{magnitude} distribution \ilya{among the three options considered here} until we have a larger number
of observations.  In Figure \ref{fig:O2-predictions}, the odds ratio
between different magnitude distributions with the same angular
distribution is much closer to unity than the odds ratio between angular
distributions.

@@

Most of our resolving power for the spin angular distribution is a
result of the fact that our ``aligned'' models cannot produce
$\chieff < 0$ (see Figure \ref{fig:chieff-distribution-models}).  If
spins are intrinsically very small, with $\ilya{\chi} \lesssim 0.3$, then it is
no longer possible to resolve the negative \ilya{effective} spin with a small number of
observations.  As noted below, however, spins observed in X-ray
binaries are typically large.  Additionally, models which do not
permit \emph{some} spins with $\chieff \gtrsim 0.1$ are ruled out by
the GW151226 observations \citep{2016PhRvL.116x1103A}.

In order to perform our analysis we need to select models for the
distribution of the spin magnitudes of stellar-mass black holes.
Observational data for such model selection is sparse.
\citet{2015PhR...548....1M} give current estimates of the spin
parameters for stellar-mass black holes, obtained using disk
reflection and/or disk continuum methods.  Most of the systems studied
are low-mass X-ray binaries rather than the high-mass X-ray binaries
that are likely to be the progenitors of double black hole binaries.
In addition, there are substantial systematic errors that can
complicate either type of analysis (see \citealt{2015PhR...548....1M}
for a discussion), and selection effects could yield a biased
distribution. Nonetheless, if we take the reported spin magnitudes as
representative then we find that there is a preference for high spins;
for example, 14 of the 19 systems with reported spins have
dimensionless spin parameters in excess of 0.5.  It is usually argued
that the masses and spin parameters of stellar-mass black holes are
unlikely to be altered significantly by accretion (low-mass donors may
not have enough mass and high-mass donors have a very short phase in
which they transfer mass; a variant of this long-standing argument was
presented by \citet{1999MNRAS.305..654K}), but see
\citet{2003MNRAS.341..385P,2015ApJ...800...17F}.  Thus the current
spin parameters probably are close to their values upon core collapse, \ilya{at least in high-mass X-ray binaries}.
However, the specific processes involved in the production of black
hole binaries from isolated binaries could alter the spin magnitude
distribution of those holes relative to the X-ray binary systems; for
example, close tidal interactions could spin up the core, or stripping
of the envelope could reduce the available angular momentum
\citep{2016MNRAS.462..844K,2017arXiv170200885Z,2017arXiv170203952H}.

@@

In summary, the angular distribution of spins in \acp{BBH} formed
through isolated binary evolution is uncertain; nevertheless, it is
more probable that spins are preferentially aligned after evolution
through this channel than not, while an isotropic distribution of
spins is the natural outcome of dynamical formation processes.
Therefore, if the current observational trend for low $\chieff$
continues with future gravitational-wave observations, it will be
possible to either confirm that \ac{BH} spins are isotropic in direction \ilya{and} sometimes large in
magnitude, yielding a strong indication of
a dynamical formation origin; or that \ac{BH} spins are overwhelmingly
small in magnitude, yielding a notable contradiction with the claimed
\ac{BH} X-ray binary spin measurements, particularly those for high-mass X-ray
binaries.

%\citet{2017CQGra..34cLT01V} also studied the possibility of distinguishing aligned and isotropic angular distributions of \ac{BBH} spins, but concluded that several hundred sources would be required to adequately separate the models.  The ``aligned'' population in that study permitted spins nearly parallel or anti-parallel to the orbital axis, i.e.\ $\hat{S}\cdot \hat{L} \simeq \pm 1$, in contrast to our aligned models; thus the aligned/isotropic discriminating power in \citet{2017CQGra..34cLT01V} relies effectively on measurements of the detailed shape of the $\chieff$ distribution and, as noted above, measurements are not very sensitive to the spin magnitude distribution that controls the shape of $\chieff$.  Note that anti-parallel spins are strongly disfavoured in models of isolated binary evolution (see, for example, \citet{Kalogera:2000}).

@@

It is possible to estimate at an order-of-magnitude level the rate at
which evidence accumulates in favour of or against the isotropic
models as more systems are detected.  Based on Figure
\ref{fig:chieff-distribution-models}, approximate the isotropic
population $\chieff$ distribution as uniform on
$\chieff \in \left[ -0.25, 0.25 \right]$ and the aligned population
$\chieff$ distribution as uniform on
$\chieff \in \left[0, 0.5\right]$.  Then the odds ratio between the
isotropic and aligned models for each event is approximately
\begin{equation}
  \label{eq:approx-odds}
  \frac{p\left( d \mid I \right)}{p\left( d \mid A \right)} \simeq
  \frac{P\left( -0.25 \leq \chieff \leq 0.25 \right)}{P\left( 0 \leq \chieff \leq 0.5 \right) },
\end{equation}
where $P\left( A \leq \chieff \leq B \right)$ is the posterior
probability \ilya{[Should really be likelihood, not posterior, right?  You are approximating the first term in the product in (B1)...]} that $\chieff$ is between $A$ and $B$.  Using our
approximations to the $\chieff$ posteriors described above, this gives
an odds ratio of $5$ in favour of the isotropic models, which is about
a factor of two smaller than the ratio in the more careful calculation
described in Section \ref{sec:O1}.  This is a satisfactory answer at
an order-of-magnitude level.

If the true distribution is isotropic and follows this simple model,
and our measurement uncertainties on $\chieff$ are $\simeq 0.1$, then
each subsequent measurement contributes on average a factor of
$\sim 3$ to the overall odds. \ilya{[This ranges from $\sim$1 to $\sim$50 (2.5 $\sigma$).  Not clear what an average of 3 means here (geometric mean, I hope?).  I am still not a huge fan of appendix B; it raises more questions than it answers, as far as I am concerned.]}  After ten additional events, then, the
odds ratio becomes $5 \times 3^{10} \simeq 3 \times 10^{5}$, or
$4.6 \sigma$, consistent with the results of the more detailed
calculation described above.  If the true distribution of spins
becomes half as wide ($\chieff \in [-0.125, 0.125]$ for isotropic and
$\chieff \in [0, 0.25]$ for aligned spins), with the same
uncertainties, then the existing odds ratio becomes $1.08$, and each
subsequent event drawn from the isotropic distribution contributes on
average a factor of $1.6$.  In this case, after 10 additional events,
the odds ratio becomes $150$, or $2.7\sigma$.  With small spin
magnitudes, our angular resolving power vanishes, as discussed in more
detail in Section \ref{sec:smallspins}.

@@

Here we extend the ``low'' model as:
%
\begin{equation}
p(\chi) \propto (1 - \chi)^{\alpha}
\label{eq:lowspinalpha}
\end{equation}
%

When $\alpha = 0$, this recovers the ``flat'' distribution, whilst $\alpha = 1$
recovers the ``low'' distribution. For higher values of $\alpha$, this distribution 
becomes more peaked towards $\ilya{\chi} = 0$.

%
\begin{figure}
\centering
\includegraphics[width=\textwidth]{../plots/sigma_v_alpha.png}
\caption{\textbf{Effect of small spins on evidence ratio of aligned against isotropic models.} The blue line shows the evidence ratio (plotted as the equivalent sigma) between a model where all systems are aligned, versus one where all systems are from an isotropic distribution as a function of the power law $\alpha$ corresponding to Equation~\ref{eq:lowspinalpha}.  The top axis shows the mean spin magnitude $\ilya{\chi}$
 \sout{that the value of} \ilya{which this $\alpha$} corresponds to. We see that for mean spin magnitudes $\lesssim 0.2$ we find no evidence for either distribution over the other.  \ilya{[Change $\bar{a}$ to $\bar{\chi}$ on upper abscissa label.]}}
\label{fig:smallspinsalpha}
\end{figure}
%

In Figure~\ref{fig:smallspinsalpha} we plot the evidence ratio of aligned to isotropic distributions (plotted as the equivalent sigma) with spin magnitudes given by this model with $\alpha$ in the range $0$ -- $6$. The top axis shows the mean spin magnitude
that value of $\alpha$ corresponds to (\ilya{e.g.}, for the ``flat'' distribution $\alpha = 0$, the mean spin magnitude is 0.5). We see that if typical BH spins are $\lesssim 0.2$ we have no evidence for one model over the other.

@@

\section{Approximations in the Gravitational Waveform}
The model waveforms used to infer the $\chieff$ of the three LIGO events
incorporate approximations to the true behaviour of the merging
systems that are expected to break down for sufficiently high
mis-aligned spins.  The effect of these approximations \ilya{on inference}  
on the parameters describing
GW150914 has been
investigated in detail \citep{2016arXiv161107531T} \ilya{[Should be an Abbott et al. ref like the others?]}.  For this source, statistical
uncertainties dominate over any waveform systematics.  Detailed
comparisons with numerical relativity computations using no
approximations to the dynamics \citep{2016PhRvD..94f4035A} also
suggest that statistical uncertainties dominate the systematics for
this system.  \citet{2016arXiv161107531T} suggests that systematics
may dominate for signals with this large SNR ($\simeq 23$) when the
source is edge-on or \ilya{has} high spins.  The other two events discussed
in this paper are at much lower SNR, with correspondingly larger
statistical uncertainties, and are probably similarly oriented and
with similarly small spins, so we do not expect systematic
uncertainties to dominate.

\sout{Given the preference of our data for a smaller spin population} We
assume here that measurements made in the future are not dominated by
systematic errors, but this assumption would need to be revisited for
high-SNR, edge-on, or high-spin sources detected in the future.  \ilya{[I thought we argued that there is no strong preference for small spin magnitudes.  We should be consistent...]}

>>>>

Figure \ref{fig:O1-posteriors} shows an approximation to the posterior
inferred on $\chieff$ for the three likely \ac{GW} detections
GW150914, GW151226 and LVT151012 from \citet{O1-BBH}.  Because samples
drawn from the posterior on $\chieff$ are not publicly released at
this time, we have approximated the posterior as a Gaussian
distribution with the same mean and 90\% credible interval as quoted
in \citet{O1-BBH}.  Since all three mergers involve black holes of
comparable masses, there is essentially no posterior support for both
black holes' $\chi \gtrsim 0.5$, in contrast to the majority of the
electromagnetically-detected stellar-mass black hole population (see
Section \ref{sec:discussion}).  The analysis here is relatively
insensitive to the precise details of the posterior distributions;
other conclusions are more sensitive.  In particular, our Gaussian
approximation does permit $\chieff = 0$ for GW151226 while the true
posterior rules this out at high confidence
\citep{2016PhRvL.116x1103A,O1-BBH}.

\begin{figure}
  \plotone{../plots/chi-eff-mock-posteriors}
  \caption{\label{fig:O1-posteriors} Approximate posteriors on
    $\chieff$ from the LIGO O1 observations in \citet{O1-BBH}.  We
    approximate the posteriors reported in \citet{O1-BBH} using
    Gaussians with the same median and 90\% credible interval as
    reported in \citet{O1-BBH}.  It is notable that none of the
    $\chieff$ posteriors extends to the regime of high spin magnitudes
    inferred from several electromagnetically-detected stellar-mass
    black hole systems (see \citet{2015PhR...548....1M} for a summary
    of such measurements).}
\end{figure}

Small values of $\chieff$ as exhibited in these systems can result
from either intrinsically small spins or larger spins whose direction
is mis-aligned with the orbital angular momentum of the binary (i.e.\
spin vectors with small $z$-components).  Mis-alignment is capable of
producing \emph{negative} values of $\chieff$, however, whereas
aligned spins will always have $\chieff \geq 0$.  This difference
provides strong discriminating power between the two angular
distributions, even without good information about the magnitude
distribution; to the extent that data favour negative $\chieff$ they
weigh heavily against aligned models.  To quantify the degree of
support for these two alternate explanations of small $\chieff$ values
in the merging \ac{BBH} population, we compared the Bayesian evidence
for various simple models of the spin population using the \ac{GW}
data set.

Each of our models for the merging \ac{BBH} spin population assumes
that the merging black holes are of equal mass (this is marginally
consistent with the three observations \citep{O1-BBH}, and the
$\chieff$ distribution is not particularly sensitive to the mass ratio
between the merging objects---see Section \ref{sec:mass-ratio}).  We
assume that the population spin distribution factorises into a
distribution for the spin magnitude $a$ and a distribution for the
spin angles.  Finally, we assume that the distribution of spins is
common to each component in a merging binary.  Choosing one of three
magnitude distributions
\begin{equation}
  \label{eq:magnitude-dists}
  p(a) = \begin{cases}
    2\left(1-a \right) & \textnormal{``low''} \\
    1 & \textnormal{``flat''} \\
    2 a & \textnormal{``high''}
  \end{cases},
\end{equation}
and pairing with either an isotropic angular distribution or a
distribution that generates perfect alignment with the positive $z$
axis yields six different models for the $\chieff$ distribution.
These models are shown in Figure \ref{fig:chieff-distribution-models}.

@@

The distributions in Eq.\ \eqref{eq:magnitude-dists} are not meant to
represent any particular physical model, but rather to capture our
uncertainty about the spin magnitude distribution, discussed in detail
in Section \ref{sec:discussion}; neither observations nor population
synthesis codes can at this point authoritatively suggest \emph{any}
particular spin distribution \citep{2015PhR...548....1M}.  Our models,
however, allow us to see how sensitive the $\chieff$ distribution is
to spin alignment given uncertainties about the spin magnitudes.

@@

Estimates of the rate of \ac{BBH} coalescences give a reasonable
chance of 10 additional \ac{BBH} detections in the next two years
\citep{O1-BBH,2016ApJ...833L...1A,2016ApJS..227...14A}.  Assuming 10
additional detections with similar observational uncertainties
\ilya{[I think it would be useful to specify (in an appendix?) what
  model we use for the observational uncertainty on $\chieff$.]}
drawn from each of our six zero-parameter models for the spin
distribution in addition to the three existing detections from O1, we
find the odds ratios shown in Figure \ref{fig:O2-predictions}.  We
find that most scenarios with an additional 10 detections allow the
simulated angular distribution to be inferred with greater than
$5\sigma$ ($2.9 \times 10^{-7}$ odds) credibility; and in the most
pessimistic case the distinction is typically
$\OTwoSigmaIsoAlignedMin\sigma$ ($\OTwoOddsIsoAlignedMin$ odds ratio).
While such future detections should permit a confident distinction
between \emph{angular} distributions, we would remain much less
certain about the \emph{magnitude} distribution among the three
options considered here until we have a larger number of observations.
In Figure \ref{fig:O2-predictions}, the odds ratio between different
magnitude distributions with the same angular distribution is much
closer to unity than the odds ratio between angular distributions.

@@

Most of our resolving power for the spin angular distribution is a
result of the fact that our ``aligned'' models cannot produce
$\chieff < 0$ (see Figure \ref{fig:chieff-distribution-models}).  If
spins are intrinsically very small, with $a \lesssim 0.3$, then it is
no longer possible to resolve the negative effective spin with a small
number of observations.  As noted below, however, spins observed in
X-ray binaries are typically large.  Additionally, models which do not
permit \emph{some} spins with $\chieff \gtrsim 0.1$ are ruled out by
the GW151226 observations \citep{2016PhRvL.116x1103A}.

In order to perform our analysis we need to select models for the
distribution of the spin magnitudes of stellar-mass black holes.
Observational data for such model selection is sparse.
\citet{2015PhR...548....1M} give current estimates of the spin
parameters for stellar-mass black holes, obtained using disk
reflection and/or disk continuum methods.  Most of the systems studied
are low-mass X-ray binaries rather than the high-mass X-ray binaries
that are likely to be the progenitors of double black hole binaries.
In addition, there are substantial systematic errors that can
complicate either type of analysis (see \citealt{2015PhR...548....1M}
for a discussion), and selection effects could yield a biased
distribution. Nonetheless, if we take the reported spin magnitudes as
representative then we find that there is a preference for high spins;
for example, 14 of the 19 systems with reported spins have
dimensionless spin parameters in excess of 0.5.  It is usually argued
that the masses and spin parameters of stellar-mass black holes are
unlikely to be altered significantly by accretion (low-mass donors may
not have enough mass and high-mass donors have a very short phase in
which they transfer mass; a variant of this long-standing argument was
presented by \citet{1999MNRAS.305..654K}), but see
\citet{2003MNRAS.341..385P,2015ApJ...800...17F}.  Thus the current
spin parameters probably are close to their values upon core collapse,
at least in high-mass X-ray binaries.  However, the specific processes
involved in the production of black hole binaries from isolated
binaries could alter the spin magnitude distribution of those holes
relative to the X-ray binary systems; for example, close tidal
interactions could spin up the core, or stripping of the envelope
could reduce the available angular momentum
\citep{2016MNRAS.462..844K,2017arXiv170200885Z,2017arXiv170203952H}.

@@

In summary, the angular distribution of spins in \acp{BBH} formed
through isolated binary evolution is uncertain; nevertheless, it is
more probable that spins are preferentially aligned after evolution
through this channel than not, while an isotropic distribution of
spins is the natural outcome of dynamical formation processes.
Therefore, if the current observational trend for low $\chieff$
continues with future gravitational-wave observations, it will be
possible to either confirm that \ac{BH} spins are isotropic in
direction and sometimes large in magnitude, yielding a strong
indication of a dynamical formation origin; or that \ac{BH} spins are
overwhelmingly small in magnitude, yielding a notable contradiction
with the claimed \ac{BH} X-ray binary spin measurements, particularly
those for high-mass X-ray binaries.

@@

It is possible to estimate at an order-of-magnitude level the rate at
which evidence accumulates in favour of or against the isotropic
models as more systems are detected.  Based on Figure
\ref{fig:chieff-distribution-models}, approximate the isotropic
population $\chieff$ distribution as uniform on
$\chieff \in \left[ -0.25, 0.25 \right]$ and the aligned population
$\chieff$ distribution as uniform on
$\chieff \in \left[0, 0.5\right]$.  Then the odds ratio between the
isotropic and aligned models for each event is approximately
\begin{equation}
  \label{eq:approx-odds}
  \frac{p\left( d \mid I \right)}{p\left( d \mid A \right)} \simeq
  \frac{P\left( -0.25 \leq \chieff \leq 0.25 \right)}{P\left( 0 \leq \chieff \leq 0.5 \right) },
\end{equation}
where $P\left( A \leq \chieff \leq B \right)$ is the posterior
probability (here used to approximate the likelihood) that $\chieff$
is between $A$ and $B$.  Using our approximations to the $\chieff$
posteriors described above, this gives an odds ratio of $5$ in favour
of the isotropic models, which is about a factor of two smaller than
the ratio in the more careful calculation described in Section
\ref{sec:O1}.  This is a satisfactory answer at an order-of-magnitude
level.

If the true distribution is isotropic and follows this simple model,
and our measurement uncertainties on $\chieff$ are $\simeq 0.1$, then
the geometric mean of each subsequent measurement's contribution to
the overall odds is $\sim 3$.  After ten additional events, then, the
odds ratio becomes $5 \times 3^{10} \simeq 3 \times 10^{5}$, or
$4.6 \sigma$, consistent with the results of the more detailed
calculation described above.  If the true distribution of spins
becomes half as wide ($\chieff \in [-0.125, 0.125]$ for isotropic and
$\chieff \in [0, 0.25]$ for aligned spins), with the same
uncertainties, then the existing odds ratio becomes $1.08$, and each
subsequent event drawn from the isotropic distribution contributes on
average a factor of $1.6$.  In this case, after 10 additional events,
the odds ratio becomes $150$, or $2.7\sigma$.  With small spin
magnitudes, our angular resolving power vanishes, as discussed in more
detail in Section \ref{sec:smallspins}.

@@

Here we extend the ``low'' model as:
%
\begin{equation}
p(a) \propto (1 - a)^{\alpha}
\label{eq:lowspinalpha}
\end{equation}
%

When $\alpha = 0$, this recovers the ``flat'' distribution, whilst
$\alpha = 1$ recovers the ``low'' distribution. For higher values of
$\alpha$, this distribution becomes more peaked towards
$a = 0$.

%
\begin{figure}
\centering
\includegraphics[width=\textwidth]{../plots/sigma_v_alpha.png}
\caption{\textbf{Effect of small spins on evidence ratio of aligned
    against isotropic models.} The blue line shows the evidence ratio
  (plotted as the equivalent sigma) between a model where all systems
  are aligned, versus one where all systems are from an isotropic
  distribution as a function of the power law $\alpha$ corresponding
  to Equation~\ref{eq:lowspinalpha}.  The top axis shows the mean spin
  magnitude $\bar{a}$ which this $\alpha$ corresponds to. We see that
  for mean spin magnitudes $\lesssim 0.2$ we find no evidence for
  either distribution over the other.}
\label{fig:smallspinsalpha}
\end{figure}
%

In Figure~\ref{fig:smallspinsalpha} we plot the evidence ratio of
aligned to isotropic distributions (plotted as the equivalent sigma)
with spin magnitudes given by this model with $\alpha$ in the range
$0$ -- $6$. The top axis shows the mean spin magnitude that value of
$\alpha$ corresponds to (e.g., for the ``flat'' distribution
$\alpha = 0$, the mean spin magnitude is 0.5). We see that if typical
BH spins are $\lesssim 0.2$ we have no evidence for one model over the
other.

@@

\section{Approximations in the Gravitational Waveform}
The model waveforms used to infer the $\chieff$ of the three LIGO
events incorporate approximations to the true behaviour of the merging
systems that are expected to break down for sufficiently high
mis-aligned spins.  The effect of these approximations on inference on
the parameters describing GW150914 has been investigated in detail
\citep{2016arXiv161107531T} \ilya{[Should be an Abbott et al. ref like
  the others?]}.  For this source, statistical uncertainties dominate
over any waveform systematics.  Detailed comparisons with numerical
relativity computations using no approximations to the dynamics
\citep{2016PhRvD..94f4035A} also suggest that statistical
uncertainties dominate the systematics for this system.
\citet{2016arXiv161107531T} suggests that systematics may dominate for
signals with this large SNR ($\simeq 23$) when the source is edge-on
or has high spins.  The other two events discussed in this paper are
at much lower SNR, with correspondingly larger statistical
uncertainties, and are probably similarly oriented and with similarly
small spins, so we do not expect systematic uncertainties to dominate.

We assume here that measurements made in the future are not dominated
by systematic errors, but this assumption would need to be revisited
for high-SNR, edge-on, or high-spin sources detected in the future.
